---
title: "Johns Hopkins COVID19 Data Project"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

The first half of this project borrows from the class lecture video. Specifically, the initial data importing and data cleaning/transformation is the same as the lecture video. The second half of the project is original. The second half contains my analyses, visualizations, and models. The sections borrowed from clas are labeled as such. 


# Bias

Before getting any further - I want to address my biases. 

I am biased toward thinking that COVID 19 was a serious issue and should be taken seriously. Hence, I would say I am biased against states/regions/groups that did not take it as seriously. I am fortunate to not have lost anyone during COVID 19, so I don't have that perspective. I also am not entirely familiar with the Northeast, South, and Midwest regions of the US. I am unfamiliar with how COVID 19 affected the rest of the world, and so I am biased toward having a US-centric perspective when it comes to the data in this project. Due to this, I chose to focus my analyses on the US, as I don't have a strong understanding of the impacts of COVID 19 globally.

# Importing Relevant Libraries - Same as Lecture Video

There are no libraries used that were not used in class. 

```{r}
library(tidyverse)
library(lubridate)
```

# Importing Data - Same as Lecture Video

The data set used for this project is a Johns Hopkins data set about COVID 19. The data will be imported from GitHub. There are five different csv files that will be used. The first four files contain data on global COVID19 cases and deaths as well as US-specific COVID19 cases and deaths. There is also a data set containing population data that will later be joined to one of the other data sets.

```{r Setting up URLs}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")

#Concatenating the start of the URL with the filenames of the desired files
urls <- str_c(url_in, file_names)

#Data set containing population data
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```

Now, the data is read in using `read_csv` and the index of the file url in the urls vector. 

```{r importing data}
US_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
US_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
uid <- read_csv(uid_lookup_url)
```

# Tidying Data - Same as Lecture Video

All of these cleaning and grouping steps are borrowed from the lecture video. 

### Tidying global data

The data set is pivoted so that "date" is a column and each row is an observation from a specific place on a specific date. Unwanted columns are removed.

```{r tidying global_cases and global_deaths data}
#Making date into a column instead of rows. Keeping Province, Country, Lat, Long as columns
global_cases <- global_cases %>% pivot_longer(cols = -c('Province/State','Country/Region','Lat', 'Long'), names_to = "date", values_to = "cases")

#Getting rid of Lat and Long columns
global_cases <- global_cases %>% select(-c('Lat', 'Long'))

#Repeating for global_deaths
global_deaths <- global_deaths %>% pivot_longer(cols = -c('Province/State','Country/Region','Lat', 'Long'), names_to = "date", values_to = "deaths") %>% select(-c('Lat', 'Long'))
```

Joining the global cases and global deaths data set. Changing the column names to make them cleaner. Changing the date column to be treated as a date variable type. 

```{r joining and cleaning global datasets}
global <- global_cases %>% full_join(global_deaths) %>% rename(Province_State = "Province/State", Country_Region = "Country/Region") %>% mutate(date = mdy(date))

#Filtering out rows with 0 cases
global <- global %>% filter(cases>0)
```
Looking at the data to see if there are issues.

```{r data inspection}
summary(global)
```

Checking if Maximum Cases seems valid by filtering for rows with numbers of cases near the maximum. Ensuring that the large number doesn't seem to be erroneous. 

```{r checking max cases}
global %>% filter(cases > 103000000) %>% head(7)
```

It appears to be correct.

### Tidying US data

Pivoting the US data. Getting rid of unwanted columns and changing date column to date data type. 

```{r tidying US_cases and US_deaths}
US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases")

US_cases <- US_cases %>% select(c(Admin2:cases)) %>% select(-c(Lat,Long_)) %>% mutate(date = mdy(date))

US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths")

US_deaths <- US_deaths %>% select(c(Admin2:deaths)) %>% select(-c(Lat,Long_)) %>% mutate(date = mdy(date))
```

Joining the two US data sets.

```{r joining the US data sets}
US <- US_cases %>% full_join(US_deaths)
```

### Changing the global data set to match the US data set

Adding columns that are in US data set to the global data set, then joining the global data set with the population data set.

```{r adjusting the global to match US}
global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ",", na.rm = TRUE, remove = FALSE)

#Filtering the population data set to only have columns we are interested in 
uid <- uid %>% select(c('Province_State','Country_Region','Population'))

#Joining the global data set with the population data set
global <- global %>% left_join(uid, by = c("Province_State", "Country_Region"))
```

### Grouping the US data

Making two new data sets with the US data. One that has the data grouped by State and date, and one that has the entire US grouped by date. 

```{r making a US_by_state data frame}
US_by_state <- US %>% group_by(Province_State, Country_Region,date) %>% 
  #Group by state, county, and date
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% 
  #Aggregate the values making cases, deaths, and pop the sum of each for each county that is grouped
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>% 
  #Adding a column that is deaths per million people
  select(Province_State,Country_Region,date,cases,deaths,deaths_per_mil,Population) %>% 
  #Selecting only certain columns
  ungroup()
```

Grouping US data by date.

```{r grouping US_by_state by date}
US_totals <- US_by_state %>% group_by(Country_Region,date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% 
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>% 
  select(Country_Region,date,cases,deaths,deaths_per_mil,Population) %>% 
  ungroup()
```

Adding columns that track new daily cases and new daily deaths.

```{r new cases and new deaths in US_totals and US_by_state}
US_by_state <- US_by_state %>% 
  mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>% 
  mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
```

Making another data frame with total cases and deaths per state. Also adding columns for cases and deaths per 1000 people. 

```{r}
US_state_totals <- US_by_state %>% group_by(Province_State) %>% 
  summarize(deaths = max(deaths), cases = max(cases), population = max(Population),
  cases_per_thou = cases * 1000 / population, deaths_per_thou = deaths * 1000 / population) %>% 
  filter(cases > 0, population > 0)
```

# Original Work

I'm choosing to focus my analyses on the US data sets specifically. 

# Summary of What Data we Have

1. Global - contains data about global cases and deaths
2. US - contains data about US cases and deaths
3. US_by_state - contains data about cases and deaths for each state by date
4. US_totals - contains total US cases and deaths by date
5. US_state_totals - contains total cases and deaths by state

```{r summary of data sets}

head(US, 1)
head(US_by_state, 1)
head(US_totals, 1)
head(US_state_totals,1)

```

The US data set has counties under "Admin2". This could be renamed to "county" to make the data clearer.
The data sets containing new cases and new deaths have NA as the first value, due to how those columns were calculated. These can be replaced with 0s, as there were 0 new cases or deaths on those dates since number of cases and deaths starts at 0.

# Further Data Cleaning

Replacing "Admin2" with "county" in the US data set. 

```{r renaming county column in US data set}

US <- US %>% rename(county = "Admin2")

```

Replacing NAs in first rows of US_totals and US_by_state new_cases and new_deaths columns with 0s.

```{r replacing NAs with 0s for new_cases and new_deaths row 1}
US_by_state$new_cases <- replace_na(US_by_state$new_cases,0)
US_by_state$new_deaths <- replace_na(US_by_state$new_deaths,0)
US_totals$new_cases <- replace_na(US_totals$new_cases,0)
US_totals$new_deaths <- replace_na(US_totals$new_deaths,0)
```

Getting rid of rows where population data is missing

```{r}
US_by_state <- US_by_state %>% filter(Population > 0)
```


# Analyses, Visualizations, and Models

## Guiding Questions

Which states had the most/fewest cases/deaths per thousand?

Does the number of cases/deaths vary by region?

Can population be used to predict cases/deaths per thousand?

## Looking at Deaths and Cases per Thousand by State

Looking at tables for the top 10 states with the most cases and deaths.

```{r}

top_state_cases <- arrange(US_state_totals, desc(cases_per_thou)) %>% select(c(Province_State, cases_per_thou, cases)) %>% head(10)

top_state_deaths <- arrange(US_state_totals, desc(deaths_per_thou)) %>% select(c(Province_State, deaths_per_thou, deaths)) %>% head(10)

print(top_state_cases)
print(top_state_deaths)
top_state_cases %>% inner_join(top_state_deaths)

```

It's interesting that the only three states that appear on both lists are Kentucky, Tennessee, and West Virginia. I have a feeling that cases and deaths per thousand are related, but they aren't so strongly related that these two lists are the exact same. It seems there were some states that had more relative cases per thousand, and not as many relative deaths per thousand. 

Doing the same thing for for the bottom 10 states with fewest cases and deaths per thousand.  

```{r}
bot_state_cases <- arrange(US_state_totals, desc(cases_per_thou)) %>% select(c(Province_State, cases_per_thou, cases,population)) %>% tail(10)

bot_state_deaths <- arrange(US_state_totals, desc(deaths_per_thou)) %>% select(c(Province_State, deaths_per_thou, deaths,population)) %>% tail(10)

print(bot_state_cases)
print(bot_state_deaths)
bot_state_cases %>% inner_join(bot_state_deaths)
```

There is more overlap among the states with the fewest cases per thousand and fewest deaths per thousand. It seems intuitive that if there are fewer cases there would be fewer deaths in those same areas. It is still interesting that the same is not entirely true for states with the most cases per thousand. 

Also, it looks like a few of these are places with smaller populations. I wonder if population size is related to cases and deaths per thousand. 

Graphing cases and deaths per thousand by state. 

```{r}
p1 <- US_state_totals %>% 
  ggplot(aes(x = reorder(Province_State,cases_per_thou, decreasing = TRUE), y = cases_per_thou, fill = Province_State)) +
  geom_bar(stat = "Identity", show.legend = FALSE) +
  xlab("State") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "Number of COVID19 Cases per Thousand People by US State", y = NULL)

p2 <- US_state_totals %>% 
  ggplot(aes(x = reorder(Province_State,deaths_per_thou, decreasing = TRUE), y = deaths_per_thou, fill = Province_State)) +
  geom_bar(stat = "Identity", show.legend = FALSE) +
  xlab("State") +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "Number of COVID19 Deaths per Thousand People by US State", y = NULL)

print(p1)
print(p2)
```


## Looking at Deaths and Cases per Region

Adding a "region" column to the US_by_state data set

```{r adding a "region" column to US_by_state}
#Defining the regions
northeast <- c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania")

midwest <- c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota")

south <- c("Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas")

west <- c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington")

US_by_state <- US_by_state %>% 
  mutate(region = ifelse(is.element(Province_State, northeast), "Northeast", 
                         ifelse(is.element(Province_State, midwest), "Midwest", 
                              ifelse(is.element(Province_State, west), "West", 
                                     ifelse(is.element(Province_State, south), "South","Other")))))

#Making sure it worked
US_by_state %>% group_by(Province_State) %>% summarize(region = max(region)) %>% select(Province_State,region) %>% arrange(,region)

#Ordering the regions
US_by_state$region <- factor(US_by_state$region, levels = c("West","Midwest", "South","Northeast", "Other"))
```

Graphing total cases per region

```{r}
US_by_state %>% group_by(Province_State,region) %>% 
  summarize(cases = max(cases)) %>% 
  group_by(region) %>% 
  summarize(cases = sum(cases)/1000000) %>% 
  ggplot(aes(x = region, y = cases, fill = region)) +
  geom_bar(stat = "Identity") +
  labs(title = "Total Number of COVID19 Cases in the US by Region") +
  xlab("Region") + ylab("Total Number of Recorded Cases (in Millions)")
```

It looks like there were more cases in the south, and fewer in the other regions. I'm curious if deaths follow a similar pattern.

Graphing total deaths per region.

```{r}
US_by_state %>% group_by(Province_State,region) %>% 
  summarize(deaths = max(deaths)) %>% 
  group_by(region) %>% 
  summarize(deaths = sum(deaths)/1000) %>% 
  ggplot(aes(x = region, y = deaths, fill = region)) +
  geom_bar(stat = "Identity") +
  labs(title = "Total Number of COVID19 Deaths in the US by Region") +
  xlab("Region") + ylab("Total Number of Recorded Deaths (in Thousands)")
```

It looks like deaths do follow a similar pattern. It's interesting that there were slightly more cases in the West than the Midwest and Northeast, but the West had fewer deaths than the Midwest, and about the same number of deaths as the Northeast.

I wonder if the number of cases and deaths per million people in each region follows a similar pattern.

Graphing number of cases per million by region.

```{r}
US_by_state %>% group_by(Province_State,region) %>% 
  summarize(cases = max(cases), Population = max(Population)) %>% 
  group_by(region) %>% 
  summarize(cases_per_mil = (sum(cases)*1000000/sum(Population))/1000) %>% 
  ggplot(aes(x = region, y = cases_per_mil, fill = region)) +
  geom_bar(stat = "Identity") +
  labs(title = "Thousands of Regional US COVID19 Cases per Million People") +
  xlab("Month") + ylab("Thousands of Cases per Million")
  
```

Doing the same for deaths.

```{r}
US_by_state %>% group_by(Province_State,region) %>% 
  summarize(deaths = max(deaths), Population = max(Population)) %>% 
  group_by(region) %>% 
  summarize(deaths_per_mil = sum(deaths)*1000000/sum(Population)) %>% 
  ggplot(aes(x = region, y = deaths_per_mil, fill = region)) +
  geom_bar(stat = "Identity") +
  labs(title = "Regional US COVID19 Deaths per Million People") +
  xlab("Month") + ylab("Deaths per Million")
```

Looking at cases and deaths per million in each region adds some perspective and reveals some interesting trends. While certain regions had more total cases and deaths than other regions, the cases per million people in each region is relatively similar. Deaths per million, however, is once again different. The Northeast and South had more deaths per million and the West and "Other" had fewer.


# Models

## Is the difference in cases/deaths per million by region significant?

Conducting an ANOVA to see if there were differences in the number of cases/deaths by region. Using p < 0.01 level of significance.

```{r}
cases_by_region <- US_by_state %>% 
  group_by(Province_State,region) %>% 
  summarize(cases = max(cases), Population = max(Population), cases_per_mil = cases*1000000/Population) 

mod1 <- aov(cases_per_mil~region, data = cases_by_region)

summary(mod1)
```

Our p-value here is 0.068, which is less than the p-value we established before conducting the test, so we fail to reject the possibility that any variations are caused by chance and not real differences in the number of cases per million in the different US regions. We can't say there were significant differences in the number of cases per million by region. 

Doing the same for deaths.

```{r}
deaths_by_region <- US_by_state %>% 
  group_by(Province_State,region) %>% 
  summarize(deaths = max(deaths), Population = max(Population), deaths_per_mil = deaths*1000000/Population) 

mod2 <- aov(deaths_per_mil~region, data = deaths_by_region)

summary(mod2)

```

Our p-value for this test is significant (less than 0.01) so we can say that there does appear to be significant differences that are unlikely due to chance. Now we need to see which regions were significantly different from each other at our 0.01 level of significance. 

```{r}
post_hoc2 <- TukeyHSD(mod2)

post_hoc2.df <- as.data.frame(post_hoc2[1])

post_hoc2.df %>% filter(region.p.adj < 0.01, ordered = TRUE)
```

Interestingly enough, the only region significantly different from the others was "Other". This region includes areas that are not states. If we conduct and ANOVA without this region, is there still a significant difference?

```{r}
deaths_by_region2 <- US_by_state %>% 
  group_by(Province_State,region) %>% 
  summarize(deaths = max(deaths), Population = max(Population), deaths_per_mil = deaths*1000000/Population) %>% 
  filter(region != "Other")

mod3 <- aov(deaths_per_mil~region, data = deaths_by_region2)

summary(mod3)
```

Now the p-value is 0.0147, which is not significant. Since we already established the p-value, we have to conclude that we fail to find evidence that deaths per million is different by region. Those differences we see could be due to chance. 

## Can population be used to predict cases/deaths per thousand?

This is similar to what was done in class, but instead of seeing if cases per thousand predicted deaths per thousand, I want to see if population can predict either of these. 

Starting with cases per thousand

```{r}
mod4 <- lm(cases_per_thou~population,data = US_state_totals)

summary(mod4)
```

It looks like population is not a significant predictor of cases per thousand (p = 0.0672).

Plotting cases per thousand by population.

```{r}
US_state_totals %>% ggplot(aes(x = population, y = cases_per_thou)) +
  geom_point(aes(x = population, y = cases_per_thou), color = "blue")+
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "US COVID 19 Cases per 1000 People by Population in Each State") +
  xlab("Population") + ylab("Cases per Thousand")
```

It is pretty evident that the data does not follow a clear pattern, and the line of best fit from our model is pretty flat.

Now doing the same thing for deaths per thousand.

```{r}
mod5 <- lm(deaths_per_thou~population,data = US_state_totals)

summary(mod5)
```

This is also not significant (p = 0.138). So population is not a significant predictor of deaths per thousand either. 

```{r}
US_state_totals %>% ggplot(aes(x = population, y = deaths_per_thou)) +
  geom_point(aes(x = population, y = deaths_per_thou), color = "blue")+
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "US COVID 19 Deaths per 1000 People by Population in Each State") +
  xlab("Population") + ylab("Deaths per Thousand")
```

It looks like there is a slightly stronger relationship between population and deaths per thousand, but still not one that is significant. 

# Conclusions

Does the number of cases/deaths vary by region?

The total number of cases and deaths does appear to vary by region, but deaths and cases per million people does not (at least for the 50 US states). The number of cases per million was not significantly different by region. The number of deaths per million in "Other" regions was significantly different, but the rest of the regions were not significantly different. 

Can population be used to predict cases/deaths per thousand?

Despite it seeming like population might have an impact on number of cases/deaths per thousand people, we were unable to find a significant effect. It doesn't appear to be the case that popluation can predict cases or deaths per thousand. 
